---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, dpi = 150,
                      fig.path = "man/figures/README-", fig.align = "center")
```

# windscape

<!-- badges: start -->
<!-- badges: end -->

Because wind disperses particles ranging from pollen, seeds, and spores to insects, pathogens, and pollutants, wind patterns shape many aspects of spatial ecology. But the variation in wind strength and direction over space and time makes it challenging to study wind's role in landscape connectivity. The **windscape** package provides a toolset for modeling the effects of time-integrated wind regimes in spatial ecology, landscape genetics, and related fields. 

While **windscape** can be used to map air flow patterns at a single moment, its core connectivity models summarize long time series of wind data into wind regimes, describing the expected connectivity averaged over many dispersal events. It is thus designed for analyzing processes that accumulate over time, such as gene flow, colonization, and range expansion. The package offers the following functionality:

| Task |  Description | Functions |
|:-----|:-------------|:----------|
| Get wind data | Download and import wind data rasters | `ncar_download()`, `ncar_land()`, `wind_series()`, `windscape_example()` |
| Model wind regimes | Summarize wind time series into | `wind_rose()`, `combine_roses()`, `weight_conductance()`, `downscale()` |
| Map windsheds | Model a site's upwind catchment area and downwind deposition shadow using *least cost path* or *random walk* approaches | `least_cost()`, `least_cost_paths()`, `random_walk()`, `ws_summarize()` |
| Compute pairwise connectivity | Model directional connectivity among sets of sites | `pairwise_least_cost()`, `pairwise_random_walk()`, `check_cell_distance()` |
| Test hypotheses | Run statistical significance tests about spatial relationships between wind and ecological outcomes | `pairwise_ratios()`, `pairwise_means()`, `mantel_test()` |
| Trace airflow | | `wind_trails()` |
| Visualize | Map spatial wind patterns in ggplot2, including instantaneous wind fields, time-integrated wind regimes, and windsheds | `geom_wind_rose()`, `geom_wind_arrow()`, `geom_wind_trail()`, `geom_wind_path()`, `scale_fill_bearing()` |
 

## Installation

Install the development version from GitHub:

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("matthewkling/windscape")
```


## Get wind data

`ncar_download()` downloads hourly wind data from NCAR's Geoscience Data Exchange, with no account needed. Datasets include ERA5 (1940 to present), CFSR (1979-2010), and CFSv2 (2011 to present). Data are clipped to your region on the server and saved as one file per month.

```{r, eval = FALSE}
library(windscape)

# a decade of 10 m altitude ERA5 winds for the western and central US, every 3rd hour
files <- ncar_download("era5", xlim = c(-120, -90), ylim = c(30, 50),
                       years = 2011:2020, time_stride = 3, dir = "~/wind_data")
```


### Visualize wind fields

A `wind_field` is a pair of rasters representing the wind conditions across a landscape at a single moment. Most **windscape** modeling analyses involve summarizing many wind fields, but we can also visualize an individual field using `geom_wind_arrow()` to draw wind vectors, or `geom_wind_trail()` to draw trails that follow the airflow, here for Hurricane Katrina on August 28, 2005. (`wind_trails()` returns the same trails as data.)

```{r katrina, fig.width = 7, fig.height = 6}
library(windscape)
library(ggplot2)

katrina <- windscape_example("wind_field")
world <- map_data("world")

ggplot(katrina, aes(x, y)) +
      geom_raster(aes(fill = speed)) +
      geom_path(data = world, aes(long, lat, group = group), color = "white", linewidth = 0.2) +
      stat_wind_trail(color = "white", hours = 6, steps = 100) +
      scale_fill_gradientn(name = "wind speed\n(m/s)", colors = c("black", "orangered")) +
      coord_quickmap(xlim = c(-99, -78), ylim = c(17, 35), expand = FALSE) +
      theme_void()
```


### Summarize wind regimes

A **wind rose** summarizes a time series of wind fields into a model of the wind regime: for each grid cell, the average conductance of wind toward each of its eight neighbors. `wind_rose()` builds one from a long series in chunks, so long records needn't fit in memory. The `trans` argument sets how wind speed translates into conductance so the rose reflects the winds that matter for dispersal, for example to model propagules that are only released in strong winds.

```{r, eval = FALSE}
rose <- wind_rose(wind_series(files), trans = 1)
```

The examples below use a small `wind_rose` object that ships with the package, built from the same region as the data downloaded above. `geom_wind_rose()` maps it as a field of glyphs, each showing the distribution of flow directions in a block of cells, colored by the direction of net flow:

```{r roses, fig.width = 8, fig.height = 5.5}
rose <- windscape_example("wind_rose")
states <- map_data("state")

ggplot(rose, aes(x, y)) +
    geom_raster(aes(alpha = speed), fill = "black") +
    geom_path(data = states, aes(long, lat, group = group), color = "black", linewidth = 0.2) +
    geom_wind_rose(res = 12, scale = 1.5) +
    coord_quickmap(xlim = ext(rose)[1:2], ylim = ext(rose)[3:4], expand = FALSE) +
    theme_void() +
    labs(alpha = "mean wind\nspeed (km/h)")
```


## Map windsheds

By analogy to watersheds, "windsheds" are the upwind catchment area where particles arriving at a site originate, or the downwind deposition area where particles released from the site end up. These two surfaces represent how wind connects a site to the surrounding landscape, and typically differ strongly from each other due to the directionality of wind patterns. windscape offers two complementary ways to model this connectivity, both of which use a `wind_rose` as input.

### Least-cost paths

Least-cost path models find the fastest route between places, giving wind travel times in hours of effective wind transport along the best route under the long-run wind regime. `least_cost()` maps travel time from a site, and `least_cost_paths()` traces a set of individual routes. Here we use them together to model *downwind* connectivity from a site in the center of the landscape.

```{r least-cost, fig.width = 8, fig.height = 5.5}
site <- cbind(-105, 40)
hours <- least_cost(rose, site, direction = "downwind")
paths <- least_cost_paths(rose, site, direction = "downwind", n = 500)

ggplot(as.data.frame(hours, xy = TRUE), aes(x, y)) +
      geom_raster(aes(fill = pmax(hours, 10))) + # floor at 10 hours for the log scale
      geom_path(data = states, aes(long, lat, group = group), color = "black", linewidth = 0.2) +
      geom_wind_path(data = paths, alpha = .5, color = "white") +
      annotate("point", site[1], site[2], color = "black", size = 3) +
      scale_fill_gradientn(name = "hours", trans = "log10", values = c(0, .5, .7, .85, 1),
                           colors = c("cyan", "dodgerblue", "purple", "red", "orange")) +
      coord_quickmap(xlim = c(-120, -90), ylim = c(30, 50), expand = FALSE) +
      theme_void()
```

### Random walks

Random walk models instead simulate particles diffusing through the wind regime, optionally with deposition at a rate set by a half-life. They capture dispersal along all routes, not just the fastest one. Here we use `random_walk()` to compare the downwind and upwind windsheds of the same site:

```{r random-walk, fig.width = 9, fig.height = 3.6}
down <- random_walk(rose, site, mode = "stream", direction = "downwind", half_life = 100)
up <- random_walk(rose, site, mode = "stream", direction = "upwind", half_life = 100)

d <- rbind(data.frame(fortify(down)[c("x", "y")], value = fortify(down)$deposition,
                      windshed = "downwind: where particles released here land"),
           data.frame(fortify(up)[c("x", "y")], value = fortify(up)$origin,
                      windshed = "upwind: where particles landing here come from"))
d$value <- d$value / ave(d$value, d$windshed, FUN = max) # relative to each windshed's peak

ggplot(d, aes(x, y)) +
      geom_raster(aes(fill = value)) +
      geom_path(data = states, aes(long, lat, group = group), color = "white", linewidth = 0.15) +
      annotate("point", site[1], site[2], color = "red", size = 2) +
      facet_wrap(~windshed) +
      scale_fill_viridis_c(name = "relative\ndensity", trans = "log10",
                           limits = c(1e-4, 1), oob = scales::squish) +
      coord_quickmap(xlim = c(-120, -90), ylim = c(30, 50), expand = FALSE) +
      theme_void() +
      theme(strip.text = element_text(margin = margin(4, 0, 4, 0)))
```

For comparing windsheds among many sites, `ws_summarize()` computes a windshed's summary statistics such as its size, its centroid, and how concentrated it is in a few compass directions.


## Estimate connectivity among sites

Many studies require pairwise connectivity estimates among a set of sites, such as sampled populations. `pairwise_least_cost()` and `pairwise_random_walk()` return matrices of wind connectivity between every pair of sites. Wind connectivity is directional, so these matrices are asymmetric: element `[i, j]` describes flow from site `i` to site `j`. The least-cost model uses each site's exact location, even for sites within the same grid cell, while the random walk model treats each site as the grid cell it falls in; `check_cell_distance()` reports how much that distorts distances among closely spaced sites.

```{r pairwise}
sites <- cbind(lon = c(-110, -105, -100, -95), lat = c(40, 42, 38, 44))

pairwise_least_cost(rose, sites) |> round() # travel time, in hours

pairwise_random_walk(rose, sites, half_life = 100) |> signif(2) # deposition density
```

These matrices can be compared with ecological data, such as genetic differentiation or gene flow, to test whether wind shapes ecological patterns. `pairwise_means()` and `pairwise_ratios()` derive matrices for testing hypotheses about the strength and the directionality of wind connectivity, and `mantel_test()` tests their relationships with other pairwise data. The package includes `birch`, a landscape genetic data set for silver birch, for trying these out.
