---
title: "Connectivity models in depth"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, out.width = "100%", dpi = 96)
```

windscape models wind connectivity from a wind rose in two ways. A **least-cost model** finds
the fastest route between places. A **random walk model** follows particles as they spread along
all routes and (potentially) settle out of the air. The [windsheds](windsheds.html) and
[pairwise connectivity](pairwise-connectivity.html) articles use both; this article explains how
they work, what their settings do, and what their limitations are, as a reference for choosing
between them and setting them up.

```{r setup}
library(windscape)
library(ggplot2)

rose <- windscape_example("wind_rose")
site <- cbind(-105, 40)

states <- map_data("state")
borders <- geom_path(data = states, aes(long, lat, group = group),
                     color = "white", linewidth = 0.15, inherit.aes = FALSE)
us <- coord_quickmap(xlim = c(-120, -90), ylim = c(30, 50), expand = FALSE)
facets <- theme(strip.text = element_text(margin = margin(4, 0, 4, 0)))
```

Both models start from the same quantity: the `wind_rose`'s **conductance** from each grid cell
toward each of its eight neighbors, giving the average rate at which wind moves material from the cell
to that neighbor. With the default `trans = 1` in `wind_rose()` and winds in m/s, conductance is in units of
1/hour, and the models' times are in hours. With other choices of `trans`, they are in relative
units, which still compare across places and directions.


## Least-cost models

### The wind graph

A least-cost model treats the grid as a network, or graph, in which each cell is linked to its
eight neighbors. Each link's conductance is the mean, over its two cells, of the rose's
conductance in the link's direction, and the cost of crossing a link is the inverse of its
conductance: the time to cover the distance between the two cell centers at the long-run average
rate that wind moves material in that direction. The cost of a route is the sum of its links'
costs, and the least-cost route between two places is the one with the lowest total cost, found
with Dijkstra's algorithm.

In a downwind graph, links point the way wind carries material, for measuring travel from a
site. An upwind graph is the same network with every link reversed, for measuring travel to a
site. The least-cost functions build the graph from a rose automatically. `wind_graph()` builds
it explicitly, which can save time when making many calls on a large grid; pass the graph in place
of the rose. `wind_graph()` also has a `wrap` argument for global grids, where the left and right
edges should be connected; the least-cost functions pass it on through `...`.

### Travel time as accessibility

Least-cost travel times are a measure of **accessibility** by wind, not a prediction of how long
particles take to arrive. Each link's conductance is a long-run average, including zero for all
the time the wind spends blowing other ways. A travel time is therefore the time to follow the
best route, including the time spent waiting, in effect, for the wind to blow roughly in the
direction of travel. Real particles riding a single favorable wind event can arrive much faster,
but most particles never arrive at all.

What travel times do capture well is the relative ease of wind transport between places, in
consistent units that compare across places and directions. This makes them well suited to
mapping windsheds and to comparing connectivity among pairs of sites.

### Eight-neighbor bias

Routes move between neighboring cells in eight directions, so a route whose bearing falls
between two neighbor directions has to zigzag, which overestimates its cost. For uniform wind on
square cells, the error is largest, about 8 percent, for routes 22.5 degrees off a grid axis. On
a longitude/latitude grid, cells narrow east-west toward the poles and the neighbor bearings
become uneven, so the maximum error grows with latitude, to roughly 18 percent at 60 degrees.
The bias is systematic rather than random, so it matters most when comparing routes in different
directions from the same place.

`pairwise_least_cost()` places sites at their actual locations rather than at cell centers, by
linking them into the graph with exact travel times through the local wind (see `?pairwise_least_cost`).
These links share the grid's directional bias, so results are consistent across distances.

### Paths form trees

`least_cost_paths()` traces the routes themselves. The fastest paths from one site to all other
places form a tree: any part of a fastest path is itself a fastest path, so once two paths meet,
they share the rest of their route back to the site. Heavily shared branches are the main
corridors of wind transport.

Because of this, the shape of the tree depends on the whole route, not just on local winds, and
small differences in travel time can determine which branch a distant point joins. Because
diagonal steps between cell centers can cross without passing through a common cell, two
branches occasionally appear to cross.


## Random walk models

### From conductance to a random walk

A random walk model moves particle mass between neighboring cells at rates set by the
conductances. Formally, the conductances define a continuous-time Markov chain, which windscape
solves by **uniformization**: it picks a time step, and converts each cell's conductances into
the probabilities of moving to each neighbor, or staying put, in one step. The default time step
is the longest possible, the time a particle spends in the cell with the highest total
conductance, which has zero probability of staying. `rw_max_step()` reports it, before the
latitude correction described below shortens it somewhat:

```{r max-step}
rw_max_step(rose)
```

The `timescale` argument, between 0 and 1, shortens the step. Mass moves only to neighboring
cells each step, so in pulse mode, which tracks a release step by step, shorter steps follow the
continuous-time process more accurately, at the cost of more iterations. Stream mode results
don't depend on the time step at all.

### Deposition and half-life

Particles leave the air at a constant rate, set by `half_life`: the time, in hours of transport,
after which half of the airborne mass remains. Mass deposited in a cell makes up its
`deposition`. With `half_life = Inf`, nothing is deposited, and particles stay airborne until
they leave the domain.

The half-life sets how far material travels, and is the main biological choice in a random walk
model. It represents everything that removes particles from the air: settling, impaction,
mortality of propagules in transit, or loss of viability. Short half-lives suit heavy propagules
or short-lived ones; long half-lives suit light, durable ones, or processes like gene flow that
accumulate over many generations of stepwise movement. Here is the downwind windshed of the same
site with three half-lives:

```{r half-life, fig.height = 2.8}
d <- do.call(rbind, lapply(c(12, 48, 192), function(h){
      w <- random_walk(rose, site, mode = "stream", half_life = h)
      data.frame(fortify(w)[c("x", "y", "deposition")], half_life = paste(h, "hours"))
}))
d$half_life <- factor(d$half_life, levels = unique(d$half_life))
d$deposition <- d$deposition / ave(d$deposition, d$half_life, FUN = max)

ggplot(d, aes(x, y)) +
      geom_raster(aes(fill = deposition)) +
      borders +
      facet_wrap(~half_life) +
      scale_fill_viridis_c(name = "relative\ndensity", trans = "log10",
                           limits = c(1e-4, 1), oob = scales::squish) +
      us + theme_void() + facets
```

Because time is in hours of transport at the rose's average wind speeds, half-lives, like
least-cost travel times, describe accessibility rather than real time aloft.

### Pulse and stream modes

`random_walk()` has two modes:

- **Pulse** mode releases particles once and follows them step by step for `iter` steps,
  recording the airborne and cumulative deposited mass at the steps in `record`. It shows how a
  release drifts and spreads over time.
- **Stream** mode releases particles continuously and returns the steady state, where release is
  balanced by deposition and loss across the domain edges. It's computed directly, with a sparse
  linear solve, rather than by simulating forward.

The two are closely linked. Particles move independently, and the same rose applies at every
step, so the steady state under continuous release equals the total, over all time, of a single
release. Stream `deposition` is therefore where a single release eventually lands, which pulse
mode approaches as it runs:

```{r modes}
pulse <- random_walk(rose, site, mode = "pulse", half_life = 48, iter = 200,
                     record = c(50, 200), density = FALSE)
stream <- random_walk(rose, site, mode = "stream", half_life = 48, density = FALSE)

terra::global(pulse$deposition, "sum", na.rm = TRUE) # cumulative deposition at steps 50 and 200
terra::global(stream$deposition, "sum", na.rm = TRUE)
```

Each step here is `r signif(iter_length(pulse), 2)` hours (`iter_length()`), so after 50 steps,
about two days, roughly half the release has been deposited, as the 48-hour half-life implies.
The stream solution gives where all of it eventually lands; the small remainder leaves the
domain. With `density = FALSE`, values are per grid cell,
so they sum to totals, as here; the default, `density = TRUE`, divides by cell area to give
values per km^2. That removes the effect of cell size on how much a cell receives, and so the
bias toward low latitudes, where cells are larger (but see Resolution below). Most windshed and
pairwise analyses use stream mode.

### Direction

A downwind walk follows particles from where they're released. An upwind walk asks the reverse
question, where particles arriving at a site come from, and is computed with the adjoint of the
downwind process rather than by reversing the wind. The two are exactly consistent: the upwind
value at a source equals the downwind value at the receptor, for any source and receptor.

Upwind stream walks with a finite half-life return **two** layers. `deposition` describes how 
likely a particle released in each cell is to be deposited at the site. `origin` give the share 
of particles deposited at the site that came from each cell, which also depends on how much each
cell releases (`source`). The [windsheds](windsheds.html) article compares them.

### Latitude correction

On a longitude/latitude grid, cells narrow east-west toward the poles. Conductances account for
this, so material drifts at the right speed at every latitude. But in a random walk on a grid,
the spread of material comes partly from the hops between cell centers themselves, and east-west
hops are shorter than north-south ones at high latitudes, which compresses east-west spread:
without correction, to roughly 0.86, 0.70, and 0.49 times north-south spread at 30, 45, and 60
degrees latitude.

`latitude_correction = TRUE`, the default, adds equal conductance toward each cell's east and
west neighbors, restoring east-west spread to what the same winds would produce on square cells,
without changing drift. The added conductance shortens the time step somewhat, which is why the
step reported by `random_walk()` is shorter than `rw_max_step()`. See the Latitude correction
section of `?random_walk` for details and limits.

### Resolution

The same effect makes random walk results depend on grid resolution: a walk's spread comes from
the hops between cells, and its rate is proportional to hop length, so walks on coarser grids
spread more (see [turbulence](#turbulence) below). Downscaling a rose with
`downscale()` narrows the spread (by 30 to 50 percent, in one test, for a factor of 4), so it
changes the model as well as the grid. Results at different resolutions aren't directly
comparable, so choose one resolution for an analysis and treat it as part of the model. Least-cost
travel times depend on resolution much less.

### Domain edges

The edges of the domain are absorbing: mass that disperses off the grid, or into `NA` cells, is
lost and never returns. Values near the edges are therefore biased low, since they receive
nothing from beyond the edge. In stream mode with a finite half-life, `random_walk()` reports the
fraction of released mass lost across the edges as a message, which gives a quick check:

```{r edge-message, message = TRUE}
w <- random_walk(rose, site, mode = "stream", half_life = 192)
```

With a 192-hour half-life, about 6 percent of this site's release leaves the domain, compared
with a few hundredths of a percent with 48 hours.

`rw_exit_prob()` maps the probability that mass released in each cell leaves the domain before
being deposited. It bounds the effect of the domain boundary on any result that's a probability
over particles released from a cell, whatever the winds beyond the edge: for example, with
`density = FALSE`, an upwind walk's `deposition` at a cell is off by at most that cell's exit
probability, and the values in a row of `pairwise_random_walk()` output are off by at most the
exit probability of the row's site, in total. Exit probabilities are highest along the downwind margins, and
grow with the half-life:

```{r exit, fig.height = 2.8}
exit <- c(rw_exit_prob(rose, half_life = 12), rw_exit_prob(rose, half_life = 48),
          rw_exit_prob(rose, half_life = 192))
names(exit) <- paste(c(12, 48, 192), "hours")

d <- as.data.frame(exit, xy = TRUE)
d <- data.frame(d[c("x", "y")], half_life = rep(names(exit), each = nrow(d)),
                exit = unlist(d[names(exit)], use.names = FALSE))
d$half_life <- factor(d$half_life, levels = names(exit))

ggplot(d, aes(x, y)) +
      geom_raster(aes(fill = exit)) +
      borders +
      facet_wrap(~half_life) +
      scale_fill_viridis_c(name = "exit\nprobability", option = "magma", limits = c(0, 1)) +
      us + theme_void() + facets
```

The usual remedy is to buffer the domain: download wind for a region extending several decay
lengths beyond the study area, until exit probabilities in the study area are acceptably small.
Exit probabilities can also screen results, by masking cells above a tolerance:

```{r exit-mask, eval = FALSE}
up <- random_walk(rose, site, mode = "stream", direction = "upwind", half_life = 48)
reliable <- terra::mask(up$deposition, rw_exit_prob(rose, half_life = 48) > 0.01, maskvalues = TRUE)
```

With `half_life = Inf`, all mass eventually leaves across the edges, so stream results measure
connectivity within the chosen domain and depend on its extent. See `?rw_exit_prob` for which
results the bound covers.

### Flux and paths

With `flux = TRUE`, a stream walk also returns the net flux of material through each cell: the
flow to each neighbor minus the flow back, as a vector. Where `residence` and `deposition`
describe the shape of a windshed, flux describes the transport that produces it. Each cell's net
outflow equals its release minus its deposition.

`random_walk_paths()` traces streamlines of this flux, giving the average routes along which
material moves. **These paths involve no randomness.** A random walk describes the spread of a
cloud of particles, each zigzagging unpredictably; the paths follow the cloud's net flow, which
is deterministic. Each path is the mean route of the material moving along it, and by default
the paths are placed so that each carries an equal share of the released material.

### Self-retention

A stream walk's value at a source cell includes the source's own release, which is often by far
the largest contribution to it. Some analyses need to leave that contribution out, for example
when predicting the occurrence of a species at each occupied cell from the propagule pressure
produced by all the other occupied cells, without each cell predicting itself.

`rw_self_retention()` computes each source cell's contribution to its own residence, per unit of
release, so it can be subtracted. Exact values take one sparse solve per cell, practical for a
set of occupied cells; the default is a fast approximation, a lower bound that counts only mass
that never leaves the cell:

```{r self}
rw_self_retention(rose, half_life = 48, cells = site)               # approximation
rw_self_retention(rose, half_life = 48, cells = site, exact = TRUE)
```

For a release raster `init`, the leave-one-out residence is `residence - init * G`, where `G` is
the self-retention at each source (with `density = FALSE`, so that both are per cell), and
leave-one-out deposition is `log(2) / half_life` times that; see `?rw_self_retention`.


## Turbulence

The air is always turbulent, but windscape models only the resolved wind: the mean flow within
each grid cell and time step of the wind data, summarized over time in the wind rose. The two
models use it differently.

A **least-cost model** is purely advective. It traces the single most favorable route through
the wind rose, with no dispersion at all, so turbulence has no natural place in it.

A **random walk** spreads material, but it's worth being precise about where that spread comes
from. Each hop, a particle moves to a neighboring cell in a direction drawn from the cell's
conductances, independently of its previous hops. The wind rose's mix of directions sets the
shape of the spread: which directions it favors, and how elongated it is. But its magnitude is
set by the length of the hops. A walk's effective diffusivity is about half the wind speed times
the hop length, so it scales with cell size. Even in perfectly steady wind, where the resolved
flow alone would carry particles together without spreading, a walk spreads them: for a steady 5 m/s wind on the
example grid, with cells about 30 km across, the walk's along-wind diffusivity is about
90,000 m^2/s, and halves each time the cell size halves. For the example rose's real winds at the
site used above, it's about 17,000 to 43,000 m^2/s, depending on direction, and also halves with
cell size.

How does this compare with the real atmosphere? At the scale of reanalysis grid cells, the
walk's diffusivity is comparable to or larger than typical estimates of horizontal dispersion by
small-scale turbulence, so adding that turbulence explicitly would change results little.
Large-scale variability is a different matter. Real winds persist, often for hours to days,
carrying particles far in one direction before turning, whereas the wind rose keeps the mix of
directions but discards their sequence, and the walk redraws its direction every hop. So the
spread of a single release in real weather can be much broader than the walk's. The random walk
is best read as a model of average connectivity under the wind regime, at a resolution that is
part of the model, rather than as a simulation of individual dispersal events.

Vertical turbulence isn't represented either. Updrafts that lift particles high into the
atmosphere drive much long-distance dispersal, and the chance of uplift depends on a particle's
settling velocity relative to the strength of turbulence. In windscape, deposition is a constant
rate set by `half_life`, the same everywhere and in every wind. Choosing wind data from a height
that matches where dispersal happens (see the [wind data](wind-data.html) article) is the main
way to account for the vertical dimension.


## Choosing a model

The two models describe the same wind rose in different ways, and suit different questions:

| | Least-cost | Random walk |
|:--|:--|:--|
| Describes | The fastest route between places | The spread of material along all routes |
| Values | Travel time (hours), or its inverse | Deposition or residence, per km^2 |
| Key setting | `trans` | `trans`, `half_life` |
| Deposition in transit | No | Yes |
| Sites | At their actual locations (`pairwise_least_cost()`) | Treated as their grid cells |
| Resolution dependence | Weak | Moderate; part of the model |
| Domain edges | Routes are confined to the domain | Mass leaving is lost; check with `rw_exit_prob()` |
| Paths | `least_cost_paths()`: fastest routes, which form a tree | `random_walk_paths()`: mean routes of material |

Least-cost models are simpler, with fewer settings, and suit questions about accessibility:
which places wind connects most readily, and in which directions. Random walk models are a closer
analog of dispersal, and suit questions about how much material moves between places, where
deposition along the way matters, at the cost of choosing a half-life and minding the domain
edges and grid resolution. Where it isn't clear which applies, running both and comparing the
results is a useful check on how much conclusions depend on the model.
