---
title: "Pairwise connectivity"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, out.width = "100%", dpi = 96)
```

Many ecological questions about wind concern a set of sites rather than a single one. Does wind
explain gene flow or genetic differentiation among populations? Do pathogens or pests spread
along the routes wind favors? Are communities more similar where wind links them strongly? Each
of these compares a pairwise matrix of wind connectivity among sites with a pairwise matrix of
ecological data.

This article covers estimating pairwise wind connectivity with windscape's two connectivity
models, placing sites on the wind grid, and testing hypotheses about the role of wind with
Mantel tests. It uses model settings such as `half_life` without explaining them in depth; the
[connectivity models](connectivity-models.html) article covers how the models work and how to
choose their settings. It ends with a worked example using `birch`, a landscape genetic data set
that ships with the package.

```{r setup}
library(windscape)
library(ggplot2)

rose <- windscape_example("wind_rose")

states <- map_data("state")
borders <- geom_path(data = states, aes(long, lat, group = group),
                     color = "gray70", linewidth = 0.15, inherit.aes = FALSE)
us <- coord_quickmap(xlim = c(-120, -90), ylim = c(30, 50), expand = FALSE)
```

We'll use twenty random sites across the example rose's domain:

```{r sites}
set.seed(2)
sites <- cbind(lon = runif(20, -114, -96), lat = runif(20, 33, 47))
```


## Least-cost connectivity

`pairwise_least_cost()` returns a matrix of least-cost wind travel times between every pair of
sites, in hours (for a rose built with `trans = 1` from winds in m/s):

```{r lc}
hours <- pairwise_least_cost(rose, sites)
round(hours[1:5, 1:5])
```

Element `[i, j]` is the travel time from site `i` to site `j`, so row `i` describes travel
downwind from site `i`, and column `j` describes travel upwind to site `j`. That's why, unlike
`least_cost()`, this function has no `direction` argument: the matrix covers both. Because wind
is directional, the matrix is asymmetric. Wind takes about `r round(hours[1, 2])` hours to carry
material from site 1, in southwestern Wyoming, to site 2, in Kansas, but about
`r round(hours[2, 1])` hours to carry it back. `least_cost_paths()` shows the two routes:

```{r lc-pair, fig.height = 4}
routes <- rbind(data.frame(least_cost_paths(rose, sites[1, , drop = FALSE], sites[2, , drop = FALSE]),
                           route = "site 1 to site 2"),
                data.frame(least_cost_paths(rose, sites[2, , drop = FALSE], sites[1, , drop = FALSE]),
                           route = "site 2 to site 1"))
routes$trail <- as.integer(factor(routes$route))

ggplot(routes, aes(x, y)) +
      borders +
      geom_wind_path(aes(color = route), linewidth = 0.8) +
      annotate("point", sites[, 1], sites[, 2], size = 0.8, color = "gray40") +
      annotate("text", sites[1:2, 1], sites[1:2, 2], label = 1:2, vjust = -0.8) +
      scale_color_manual(values = c("darkorange", "dodgerblue")) +
      us + theme_void()
```

Travel times are a measure of accessibility, small for well-connected pairs. `rate = TRUE`
returns their inverse, a flow rate, which is large for well-connected pairs, matching the
orientation of random walk connectivity.


## Random walk connectivity

`pairwise_random_walk()` estimates connectivity with a stream-mode random walk released from
each site in turn. By default, element `[i, j]` is the probability density (per km^2) that a
particle released at site `i` is deposited at site `j`:

```{r rw}
deposition <- pairwise_random_walk(rose, sites, half_life = 48)
signif(deposition[1:5, 1:5], 2)
```

A half-life is required under the default `value = "deposition"`, since without it nothing is 
deposited. `value = "residence"` instead gives the time particles from site `i` spend airborne 
over site `j`, which works without a half-life. Values are per km^2 of the destination's grid 
cell (`density = TRUE`), because a larger cell catches more of the particles passing over it; 
on a longitude/latitude grid, cell area shrinks toward the poles, so per-cell values would favor 
low-latitude destinations.

The diagonal holds each site's self-connectivity: its own release deposited in its own cell.
It's usually far larger than the other values, and the tests below ignore it.

Deposition spans many orders of magnitude, from relatively large values for close downwind
neighbors to vanishingly small ones for distant upwind neighbors, so it's usually analyzed on a
log scale.


## Comparing the models

The two models measure different things: the speed of the best route, versus the share of
material that arrives along all routes, net of deposition on the way. Plotting one against the
other shows how they relate for these sites:

```{r compare, fig.height = 4}
off <- row(hours) != col(hours) # pairs of distinct sites
d <- data.frame(rate = 1 / hours[off], deposition = deposition[off])

ggplot(d, aes(rate, deposition)) +
      geom_point(alpha = 0.5) +
      scale_x_log10() + scale_y_log10() +
      labs(x = "least-cost flow rate (1 / hours)", y = "random walk deposition (per km^2)") +
      theme_minimal()

cor(d$rate, d$deposition, method = "spearman")
```

The models largely agree on which pairs are well connected, but not on how connectivity scales.
Among poorly connected pairs, random walk connectivity falls off far faster than flow rate, so it
concentrates on near, downwind neighbors. Reaching a distant upwind site requires material to
spread against or across the prevailing wind, which becomes exponentially unlikely with
distance, while a least-cost route can always wait out the wind. Deposition along the way adds to
this, but only modestly: a longer half-life flattens the curve only slightly. The models also
diverge where the fastest route is narrow or roundabout, so that little material actually
follows it.

Least-cost connectivity suits questions about accessibility, and handles sites that are close
together (see below). Random walk connectivity suits questions about how much material is
exchanged, such as dispersal or gene flow, especially when deposition along the way matters.
Where the choice isn't clear, testing both is informative.


## Placing sites on the grid

Both models work on the wind rose's grid, but they place sites on it differently, which matters
when sites are only a few grid cells apart.

`pairwise_least_cost()` places sites at their actual locations. It adds each site to the wind
graph as a node of its own, linked to the centers of nearby grid cells, and to other nearby
sites, by travel times computed exactly for the local wind. It therefore distinguishes sites
that are close together, even within the same grid cell. With `snap = TRUE`, sites are instead
moved to the centers of their cells, as in other least-cost tools. Here are three sites, the
first two in the same grid cell:

```{r snap}
near <- cbind(c(-104.95, -104.75, -104.3), c(40.1, 40.25, 40.15))
terra::cellFromXY(rose, near)

round(pairwise_least_cost(rose, near), 1)
round(pairwise_least_cost(rose, near, snap = TRUE), 1)
```

Snapped, the first two sites are zero hours apart, with identical travel times to and from the
third site. Placed at their actual locations, they get distinct travel times, which change
smoothly as sites move.

A random walk moves particles from cell to cell, so `pairwise_random_walk()` treats each site as
its grid cell: sites in the same cell get identical rows and columns, and the distances and
directions among sites a few cells apart are distorted. `check_cell_distance()` reports how
much the grid distorts the distances among a set of sites:

```{r check, message = TRUE}
check_cell_distance(rose, sites)
```

None of the twenty sites share a cell, and most discrepancies are a few percent, so this grid
suits them. Where many pairs are affected, a finer grid separates nearby sites, from finer wind
data or from `downscale()`, which interpolates a wind rose onto a finer grid. But a random
walk's spread depends on cell size, so a finer grid also changes the model: in one test,
downscaling by a factor of 4 narrowed a walk's spread by 30 to 50 percent. Choose one resolution
for an analysis and treat it as part of the model specification; see `?downscale`.


## Testing hypotheses

Pairwise wind connectivity can be compared with pairwise ecological data to test whether wind
shapes ecological patterns. Kling and Ackerly (2021) framed several kinds of hypotheses, each
using a different summary of the connectivity matrix:

- **Flow**: is directional wind connectivity related to directional ecological flow, such as
  gene flow or the spread of a pathogen? Compare the asymmetric matrices directly.
- **Isolation**: are pairs of sites that are poorly connected by wind, in both directions, more
  different, for example genetically? `pairwise_means()` averages the two directions into a
  symmetric matrix of overall connectivity, isolating the effects *strength*.
- **Asymmetry**: are directional imbalances in wind connectivity related to imbalances in
  ecological flow? `pairwise_ratios()` converts a matrix into log ratios of the two directions,
  `log(x[i, j] / x[j, i])`, isolating the effects of *direction*.

`pairwise_ratios()` also accepts a vector of site attributes, returning the log ratio of each
pair's values. This supports site-level hypotheses such as whether sites that receive more wind
than they send, being downwind of the others, have higher genetic diversity (see the worked
example below).

### Mantel tests

The values in a pairwise matrix aren't independent, since each site contributes to many pairs, so
ordinary correlation tests overstate significance. `mantel_test()` assesses significance by
permuting sites, rearranging the rows and columns of one matrix together. Unlike most Mantel
implementations, it handles asymmetric matrices and partial tests with several control matrices
(`z`), such as geographic distance or environmental difference. Diagonals are ignored.

To demonstrate, we'll simulate gene flow that partly reflects wind connectivity, and genetic
differentiation that is low where gene flow is high:

```{r simulate}
set.seed(1)
n <- nrow(sites)
log_flow <- log(deposition) + rnorm(n^2, sd = 8) # wind signal plus noise
gene_flow <- exp(log_flow)
gene_diff <- pairwise_means(matrix(rnorm(n^2, sd = 4), n)) - pairwise_means(log_flow)

distance <- point_distance(sites) # km
```

Now we test each hypothesis, controlling for geographic distance in the flow and isolation
tests, since nearby sites tend to be both well connected by wind and ecologically similar. 
(Distance control is unneeded for the asymmetry test, since `pairwise_ratios()` generates
reciprocally symmetrical matrices that, by construction, are uncorrelated with symmetric covariates 
like distance.)

```{r mantel}
tests <- list(
      flow = mantel_test(log(deposition), log(gene_flow), z = list(distance)),
      isolation = mantel_test(log(pairwise_means(hours)), gene_diff, z = list(distance)),
      asymmetry = mantel_test(pairwise_ratios(deposition), pairwise_ratios(gene_flow)))

sapply(tests, function(x) round(c(stat = x$stat, p.value = x$p.value), 3))
```

All three relationships are detected, as they should be for data simulated this way. A few
points matter for real analyses:

- **Orientation.** Travel times are small for well-connected pairs, and deposition and flow rates
  are large. The isolation test above expects a positive correlation, because pairs that take
  longer to reach each other should be more differentiated; with deposition it would be negative.
  Using `rate = TRUE` for least-cost matrices puts both models in the same orientation.
- **Transformations.** Connectivity values are strongly skewed, and deposition spans many orders
  of magnitude, so log transformations usually make relationships closer to linear. Mantel tests
  use Pearson correlation by default; `method = "spearman"` tests rank correlation instead.
- **Power.** Mantel tests have limited power with few sites. The number of permutations
  (`nperm`, default 999) sets the resolution of p-values: a p-value of 0, as above, means no
  permutation matched the observed correlation, and is better reported as p < 0.001.
- **Model settings.** Results can depend on the model and its settings, such as `half_life` and
  `trans`. Choose these from the biology of the system where possible, and report how
  sensitive results are to them, rather than searching for the settings that fit best.


## Worked example: silver birch

The `birch` data set holds landscape genetic data for silver birch (*Betula pendula*) from 29
populations across Eurasia (Tsuda et al. 2017): site coordinates, allelic richness at each site
(`div`), pairwise relative gene flow (`mig`), and pairwise genetic differentiation (`fst`).

```{r birch}
str(birch, max.level = 1, give.attr = FALSE)
```

The sites span Europe and Siberia, outside the example wind data, so the code below starts by
downloading wind for the region. It isn't run here, since the download is large.

```{r birch-wind, eval = FALSE}
files <- download_wind_data("era5", xlim = c(-15, 100), ylim = c(35, 72),
                       years = 2011:2020, hours = c(0, 6, 12, 18), dir = "~/wind_data")
rose <- wind_rose(wind_series(files), trans = 1)
```

With the rose in hand, the analysis follows the steps above. Here we use least-cost flow rates,
so that every matrix is oriented with large values meaning strong connectivity:

```{r birch-analysis, eval = FALSE}
sites <- birch$sites
wind <- pairwise_least_cost(rose, sites, rate = TRUE)
distance <- point_distance(sites)

tests <- list(
      # does gene flow follow wind?
      flow = mantel_test(log(wind), log(birch$mig), z = list(distance)),
      # are populations poorly connected by wind more differentiated?
      isolation = mantel_test(log(pairwise_means(wind)), birch$fst, z = list(distance)),
      # is gene flow stronger in the direction wind favors?
      asymmetry = mantel_test(pairwise_ratios(wind), pairwise_ratios(birch$mig)),
      # are populations downwind of others more diverse?
      diversity = mantel_test(pairwise_ratios(wind), pairwise_ratios(birch$div)))

sapply(tests, function(x) round(c(stat = x$stat, p.value = x$p.value), 3))
```

The flow and asymmetry tests expect positive correlations, and the isolation test a negative
one. For the diversity test, `pairwise_ratios(wind)[i, j]` is large when wind favors flow from
site `i` to site `j`, putting `j` downwind, so if downwind sites are more diverse,
`pairwise_ratios(birch$div)[i, j]`, the log of `div[i] / div[j]`, should be small: the test
expects a negative correlation. Swapping in `pairwise_random_walk()` tests the same hypotheses
with random walk connectivity.


## References

- Kling, M. M., and D. D. Ackerly. 2021. Global wind patterns shape genetic differentiation,
  asymmetric gene flow, and genetic diversity in trees. *Proceedings of the National Academy of
  Sciences* 118: e2017317118.
- Tsuda, Y., V. Semerikov, F. Sebastiani, G. G. Vendramin, and M. Lascoux. 2017. Multispecies
  genetic structure and hybridization in the *Betula* genus across Eurasia. *Molecular Ecology*
  26: 589-605.
